---
title: "Regime shifts paper"
date: "5/13/2020"
output: pdf_document
---

# Data preparation

Imports required packages for the analysis.
```{r, message=FALSE}
require(tidyverse)
require(zoo) # for rolling avg function
```


The SRJS algae and ancillary environmental data comes in long format, which is imported below. 
```{r}
algae_master <- read.csv(file = "../Data/algae_spp_all.csv", 
                         header = T, 
                         sep = ",")

ancillary_master <- read.csv(file = "../Data/Master_SRJS_Query.csv", 
                             header = T, 
                             sep = ",") %>%
  mutate(TP.ug.L = TP..mg.L.*1000) %>%
  group_by(Site_ID) %>%
  mutate(TP.rolling = rollapply(data = TP.ug.L,
                                width = 4, # 4 specifies 6-mo window
                                FUN = mean,
                                partial = TRUE,
                                na.rm = TRUE,
                                fill = NA,
                                align = "right"),
         TN.rolling = rollapply(data = TN..mg.L.,
                                width = 4, # 4 specifies 6-mo window
                                FUN = mean,
                                partial = TRUE,
                                na.rm = TRUE,
                                fill = NA,
                                align = "right")) %>%
  #################################################################
  # Excludes sampling events we do not have algal taxonomy data for
  # WILL NEED TO CHANGE BASED ON ANALYSIS!
  #################################################################
  filter(Event != 0) %>%
  filter(Event != 10) %>%
  filter(Event != 11.5)
```

The environmental data frame houses more variables that are needed for this analysis, so I compute summary stats on the environmental variables of interest, as well as convert total phosphorus concentrations to $\mu$g/L for plotting purposes.

```{r, message=FALSE}
query_summary <- ancillary_master %>% 
  group_by(Site_ID) %>% 
  select(Q_CFS:TN.rolling) %>% # select which ENV variables to run summarise_each on
  summarise_all(funs(mean(., na.rm = TRUE), 
                      min(., na.rm = TRUE), 
                      max(., na.rm = TRUE),
                      sd(., na.rm = TRUE),
                      median(., na.rm = TRUE),
                      IQR(., na.rm = TRUE)))


query_summary$Site_ID <- reorder(query_summary$Site_ID, query_summary$TP..mg.L._mean)
```

Below gets the algae taxonomy data into wide format.

```{r}
alg_wide <- algae_master %>%
  select(Event, Site_ID, ShortName, Biovolume_mm3_m2) %>%
  group_by(Site_ID, Event, ShortName) %>%
  summarise(Biovolume_mm3_m2 = sum(Biovolume_mm3_m2)) %>%
  spread(ShortName, Biovolume_mm3_m2, fill = 0)
```

