---
title: "Analysis"
output: html_document
date: "2024-12-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(mgcv)
library(marginaleffects)
library(TITAN2)

```


```{r}


alg_df <- read_csv(here("data/alg_df_resolved.csv"))
alg_names <- alg_df|>distinct(ShortName, Species)

alg_wide <- alg_df|>
  select(Event, Site_ID, ShortName, Biovolume_mm3_m2)|>
  group_by(Site_ID, Event, ShortName)|>
  summarise(Biovolume_mm3_m2 = sum(Biovolume_mm3_m2, na.rm = TRUE))|>ungroup()|>
  pivot_wider(names_from = ShortName, values_from = Biovolume_mm3_m2, values_fill = 0)

clad_df<-alg_wide|>
  pivot_longer(-c(Site_ID, Event), names_to = "ShortName", values_to = "Biovolume_mm3_m2")|>
  left_join(alg_names)|>
  filter(str_detect(Species, "Clad"))
  


inv_df <- read_csv(here("data/SRJS_inverts.csv"))

sites <- read_csv(here("data/SRJS_sites.csv"))



env_df1 <- read_csv(here("data/Master_SRJS_Query.csv"))|>
  filter(Event != 0)|>left_join(clad_df)|>
  mutate(TP_ug.L = 1000*`TP (mg/L)`,
         TN_ug.L = 1000*`TN (mg/L)`,
         TNTP_molar = TN_ug.L/TP_ug.L *31/14)|>
  rename(Chla_mg.m2 = `CHLA mg/m2`, NP_bulk = `NP bulk`,clad_biovol = "Biovolume_mm3_m2")

env_df <- env_df1|>
  group_by(Site_ID)|>
  summarize(across(c(NP_bulk, TP_ug.L, TN_ug.L, Chla_mg.m2, clad_biovol), ~median(., na.rm = TRUE)))|>
  mutate(TNTP_molar = TN_ug.L/TP_ug.L *31/14)|>
  rename()
```
Exploratory Viz
```{r}
ggplot(data = env_df, aes(x = TP_ug.L))+geom_histogram()
ggplot(data = env_df, aes(x = TN_ug.L))+geom_histogram()
ggplot(data = env_df, aes(x = TNTP_molar))+geom_histogram()+scale_x_continuous(breaks = c(0,10,25,50,100,150,200))

ggplot(data = env_df|>mutate(Site_ID = fct_reorder(Site_ID, TP_ug.L, desc)), aes(x = Site_ID, y = TP_ug.L))+geom_point()+theme(axis.text = element_text(angle = 90))+
  scale_y_continuous(breaks = c(0,5,10,20,30,60,90,110))

ggplot(data = env_df1|>mutate(Site_ID = fct_reorder(Site_ID, TP_ug.L, ~desc(mean(., na.rm = TRUE)))), aes(x = Site_ID, y = TP_ug.L))+geom_point()+theme(axis.text = element_text(angle = 90))+scale_y_log10(breaks =  c(2.5, 5,7.5,10,30,100,300))
  scale_y_continuous(breaks = c(0,5,10,20,30,60,90,110))

ggplot(data = env_df|>mutate(Site_ID = fct_reorder(Site_ID, TN_ug.L, desc)), aes(x = Site_ID, y = TN_ug.L))+geom_point()+theme(axis.text = element_text(angle = 90))+
  scale_y_continuous(breaks = c(0,250,1000,2000,3000,4000,5000))


ggplot(data = env_df, aes(x = TNTP_molar, y=NP_bulk, color = clad_biovol, size = clad_biovol))+geom_point()+scale_x_continuous(breaks = c(0,10,25,50,100,150,200))
ggplot(data = env_df, aes(x = TP_ug.L, y = TN_ug.L))+geom_point()+
  #scale_y_log10()+scale_x_log10()+
  geom_smooth(method = "lm")+geom_abline(slope = 1)

ggplot(data = env_df, aes(x = TP_ug.L, y = TNTP_molar))+geom_point()+geom_smooth() # most basic assumption would be that as TP increases, N:P decreases, but that's OPPOSITE of what's going on at the low end of the TP gradient - as TP increases, TN is increasing even more quickly

ggplot(data = env_df, aes(x = TP_ug.L, y = Chla_mg.m2))+geom_point()+geom_smooth(method = "gam")
ggplot(data = env_df, aes(x = TN_ug.L, y = Chla_mg.m2))+geom_point()+geom_smooth(method = "gam")
ggplot(data = env_df, aes(x = NP_bulk, y = Chla_mg.m2, color = clad_biovol, size = clad_biovol))+geom_point()+geom_smooth(method = "gam")
ggplot(data = env_df, aes(x = TNTP_molar, y = Chla_mg.m2))+geom_point()+geom_smooth(method = "gam")

ggplot(data = env_df, aes(x = clad_biovol, y = Chla_mg.m2))+geom_point()+geom_smooth(method = "gam")
ggplot(data = env_df, aes(x = log(clad_biovol), y = Chla_mg.m2))+geom_point()+geom_smooth(method = "gam")


ggplot(data = env_df, aes(x = TN_ug.L, y = clad_biovol, size = TP_ug.L))+geom_point()+geom_smooth(method = "gam")
ggplot(data = env_df1, aes(x = TP_ug.L, y = clad_biovol, color = TEMP))+geom_point()+geom_smooth(method = "gam")+
  scale_x_continuous(limits = c(0,150))+
  facet_wrap(.~Event, scales = "free_y")

```

Visualize richness trends
```{r}
rich_df <- inv_df|>
  rowwise()|>
  mutate(Richness = sum(c_across(-c(Site_ID, Event)) > 0))|>
  select(Site_ID, Event, Richness)|>
  left_join(env_df)

ggplot(data = rich_df, aes(x = log(TP_ug.L), y = Richness, group = Event))+geom_point()+geom_smooth()+
  facet_wrap(.~Event)

ggplot(data = rich_df, aes(x = log(clad_biovol+10), y = Richness, group = Event))+geom_point()+geom_smooth()+
  facet_wrap(.~Event)

```

# Site Map

LOTS of site details/summaries can be found in Cook, Stephen C., et al. 2018. Freshwater eutrophication drives sharp reductions in temporal beta diversity. Ecology 99:47-56. https://doi.org/10.1002/ecy.2069.

```{r}


```

# GAM
Want to test relationships:
1) Chla ~ TP + TN
2) Chla ~ Cladophora
3) Cladophora ~ TP + TN
```{r model fitting}

m_n <- gam(data=env_df, Chla_mg.m2 ~ s(TN_ug.L), family = Gamma(link = "log"), method = "REML")
m_p <- gam(data=env_df, Chla_mg.m2 ~ s(TP_ug.L), family = Gamma(link = "log"), method = "REML")
m_np <- gam(data=env_df, Chla_mg.m2 ~ s(TP_ug.L) +s(TN_ug.L), family = Gamma(link = "log"), method = "REML")

m_np2 <- gam(data=env_df, Chla_mg.m2 ~  s(TP_ug.L)+s(TN_ug.L) + s(TNTP_molar), family = Gamma(link = "log"), method = "REML")

m_np3 <- gam(data=env_df, Chla_mg.m2 ~ s(TNTP_molar), family = Gamma(link = "log"), method = "REML")


m_np_te <- gam(data=env_df, Chla_mg.m2 ~ te(TP_ug.L, TN_ug.L), family = Gamma(link = "log"), method = "REML")

newdata_df <- tibble(TN_ug.L = c(rep(100, times = 100),rep(200, times = 100),rep(1000, times = 100),rep(2000, times = 100)),
                     TP_ug.L = rep(seq(5,125, length.out = 100), times = 4),
                     TNTP_molar = rep(seq(20,150, length.out = 100), times = 4))


plot_predictions(m_n, condition = "TN_ug.L", points = 0.5)
plot_predictions(m_p, condition = "TP_ug.L", points = 0.5)
plot_predictions(m_np, condition = "TN_ug.L", points = 0.5)
plot_predictions(m_np, condition = "TP_ug.L", points = 0.5)

plot_predictions(m_np3, condition = "TNTP_molar", points = 0.5)

m_np_preds<-predictions(m_np, newdata = newdata_df)|>
  mutate(TN_ug.L = as_factor(TN_ug.L))

ggplot(data = m_np_preds)+
  geom_line(data = m_np_preds, aes(x = TP_ug.L, y = estimate, color = TN_ug.L))+
  geom_line(data = m_np_preds,aes(x = TP_ug.L, y = conf.high, color = TN_ug.L), linetype = "dashed")+
  geom_line(data = m_np_preds,aes(x = TP_ug.L, y = conf.low, color = TN_ug.L), linetype = "dashed")+
  geom_point(data = env_df, aes(x = TP_ug.L, y = Chla_mg.m2))+
  #facet_wrap(.~TN_ug.L)+
  theme_bw()


plot_predictions(m_np, condition = "TN_ug.L", points = 0.5)
plot_predictions(m_np, condition = "TP_ug.L", points = 0.5)

```

```{r model fitting}
env_df2 <- env_df|>
  mutate(clad_biovol = ifelse(clad_biovol == 0, clad_biovol+1, clad_biovol))
m_n <- gam(data=env_df2, clad_biovol ~ s(TN_ug.L), family = Gamma(link = "log"), method = "REML")
m_p <- gam(data=env_df2, clad_biovol ~ s(TP_ug.L), family = Gamma(link = "log"), method = "REML")
m_np <- gam(data=env_df2, clad_biovol ~ TP_ug.L + TN_ug.L + s(TP_ug.L) +s(TN_ug.L), family = Gamma(link = "log"), method = "REML")


m_np_te <- gam(data=env_df2, clad_biovol ~ te(TP_ug.L, TN_ug.L), family = Gamma(link = "log"), method = "REML")

newdata_df <- tibble(TN_ug.L = c(rep(100, times = 100),rep(200, times = 100),rep(1000, times = 100),rep(2000, times = 100)),
                     TP_ug.L = rep(seq(5,125, length.out = 100), times = 4))


plot_predictions(m_n, condition = "TN_ug.L", points = 0.5)
plot_predictions(m_p, condition = "TP_ug.L", points = 0.5)
plot_predictions(m_np, condition = "TN_ug.L", points = 0.5)
plot_predictions(m_np, condition = "TP_ug.L", points = 0.5)

m_np_preds<-predictions(m_np, newdata = newdata_df)|>
  mutate(TN_ug.L = as_factor(TN_ug.L))

ggplot(data = m_np_preds)+
  geom_line(data = m_np_preds, aes(x = TP_ug.L, y = estimate, color = TN_ug.L))+
  geom_line(data = m_np_preds,aes(x = TP_ug.L, y = conf.high, color = TN_ug.L), linetype = "dashed")+
  geom_line(data = m_np_preds,aes(x = TP_ug.L, y = conf.low, color = TN_ug.L), linetype = "dashed")+
  geom_point(data = env_df2, aes(x = TP_ug.L, y = clad_biovol))+
  #facet_wrap(.~TN_ug.L)+
  theme_bw()


plot_predictions(m_np, condition = "TN_ug.L", points = 0.5)
plot_predictions(m_np, condition = "TP_ug.L", points = 0.5)

```


# Algal TITAN


# Invert 

```{r prep data}
inv_com_df <- inv_df|>
  pivot_longer(-c(1,2), names_to = "taxon", values_to = "count")|>
  mutate(present = ifelse(count >0, 1,0))|>
  group_by(taxon)|>
  mutate(taxon_present_sum = sum(present))|>
  filter(taxon_present_sum>5)|> # drop taxa that are present less than 5 times
  select(-present,-taxon_present_sum)|>
  pivot_wider(values_from = count, names_from = taxon)|> #pivot wider again for species/site matrix
  left_join(env_df1|>select(Site_ID, Event, TP_ug.L, TN_ug.L, clad_biovol))

inv_com_df2 <- inv_df|>
  filter(!Site_ID %in% c("LLEE1", "LEE1", "MTFK1", "COVE"))|>
  pivot_longer(-c(1,2), names_to = "taxon", values_to = "count")|>
  mutate(present = ifelse(count >0, 1,0))|>
  group_by(taxon)|>
  mutate(taxon_present_sum = sum(present))|>
  filter(taxon_present_sum>5)|> # drop taxa that are present less than 5 times
  select(-present,-taxon_present_sum)|>
  pivot_wider(values_from = count, names_from = taxon)|> #pivot wider again for species/site matrix
  left_join(env_df1|>select(Site_ID, Event, TP_ug.L, TN_ug.L, clad_biovol))
```
## NMDS
Are there taxa uniquely associated with Boston mountains sites? Should we drop those from the TITAN analysis? They ar

```{r}
com <- inv_com_df|>select(-Site_ID, -Event,-TP_ug.L, -TN_ug.L, -clad_biovol)
sites_tib <- invert_df2|>select(site_date)

```


```{r classic distance-based ordination - NMDS}



mds_com <- metaMDS(com,
                  distance = "bray", k = 2, trymax = 40)
#default sqrt trans, wisconsin double standardization
#converged with k = 2, stresses 0.27
# k=3 stress ~0.2

com_env <- inv_com_df|>select(Site_ID,Event,TN_ug.L,TP_ug.L)


# envfit

com_envfit <- envfit(mds_com ~ TN_ug.L+TP_ug.L, choices=c(1:3),data = com_env, 
                    permutations = how(nperm = 999), strata = com_env$Event)


#Plotting

theme_mds <- theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.text = element_text(size = 16), axis.text = element_text(face = "bold", size = 14), axis.title = element_text(size = 18, face = "bold"), plot.title = element_text(size = 18, face = "bold"))

## NMDS1 vs 2
env_arrows<-(as.data.frame(scores(com_envfit, "vectors")) * ordiArrowMul(com_envfit))|>as_tibble(rownames = "Env_Variable")

species_scores <- scores(mds_com)$species|>as_tibble(rownames = "Chem_Var")
site_scores <- scores(mds_com)$sites|>as_tibble(rownames = "Site")|>bind_cols(inv_df|>select(Site_ID))

mds_com.p1 <- ggplot()+
 # ggrepel::geom_text_repel(data = species_scores, aes(x = NMDS1, y = NMDS2, label = Chem_Var), color = "red", max.overlaps = 15)+
  ggrepel::geom_text_repel(data = site_scores, aes(x = NMDS1, y = NMDS2, label = Site_ID), color = "red", max.overlaps = 15)+
  ggrepel::geom_text_repel(data = env_arrows, aes(x = NMDS1, y = NMDS2, label = Env_Variable), max.overlaps = 15)+
 geom_segment(data = env_arrows, aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2), color = "black", arrow = arrow(length = unit(0.3,"cm")), linewidth = 1)+
  theme_mds+
  theme(legend.position = c(0.14,0.88), legend.title =element_blank(),legend.box.background = element_rect(colour = "black"))

mds_com.p1

```

## TITAN





```{r run titan}
inv_com_df_taxa <- inv_com_df|>select(-Site_ID, -Event,-TP_ug.L, -TN_ug.L, -clad_biovol)

inv_com_df_taxa2 <- inv_com_df2|>select(-Site_ID, -Event,-TP_ug.L, -TN_ug.L, -clad_biovol)


inv_com_dfclad<-inv_com_df|>filter(!is.na(clad_biovol))|>select(-Site_ID, -Event,-TP_ug.L, -TN_ug.L, -clad_biovol)
inv_cladenv_df <-inv_com_df|>filter(!is.na(clad_biovol))

inv_com_dfclad2 <-inv_com_df2|>filter(!is.na(clad_biovol))|>select(-Site_ID, -Event,-TP_ug.L, -TN_ug.L, -clad_biovol)
inv_cladenv_df2 <-inv_com_df2|>filter(!is.na(clad_biovol))

# TITAN runs

titan_tp <- titan(env = inv_com_df$TP_ug.L, txa = inv_com_df_taxa, ncpus = 14,nBoot = 999, numPerm = 999)
saveRDS(titan_tp, here("data/titan_tp.Rds"))
titan_tn <- titan(env = inv_com_df$TN_ug.L, txa = inv_com_df_taxa, ncpus = 14,nBoot = 999, numPerm = 999)
saveRDS(titan_tn, here("data/titan_tn.Rds"))

titan_tp2 <- titan(env = inv_com_df2$TP_ug.L, txa = inv_com_df_taxa2, ncpus = 14,nBoot = 999, numPerm = 999)
saveRDS(titan_tp2, here("data/titan_tp2.Rds"))
titan_tn2 <- titan(env = inv_com_df2$TN_ug.L, txa = inv_com_df_taxa2, ncpus = 14,nBoot = 999, numPerm = 999)
saveRDS(titan_tn2, here("data/titan_tn2.Rds"))

titan_clad <- titan(env = inv_cladenv_df$clad_biovol, txa = inv_com_dfclad, ncpus = 14,nBoot = 999, numPerm = 999)
saveRDS(titan_clad, here("data/titan_clad.Rds"))

titan_clad2 <- titan(env = inv_cladenv_df2$clad_biovol, txa = inv_com_dfclad2, ncpus = 14,nBoot = 999, numPerm = 999)
saveRDS(titan_clad2, here("data/titan_clad2.Rds"))
#about 80 minutes for this titan run 

```

```{r}

titan_tp_df <- titan_tp$sppmax|>as_tibble(rownames = "taxon")
head(titan_tp_df)

plot_taxa(titan_tp,
xlabel = "TP")
plot_taxa(titan_tp2,
xlabel = "TP")
plot_taxa(titan_tn,
xlabel = "TN")
plot_taxa(titan_clad2,
xlabel = "clad")

plot_sumz_density(titan_tp, ribbon = FALSE, sumz = TRUE, change_points = TRUE)
plot_sumz_density(titan_tp2, ribbon = FALSE, sumz = TRUE, change_points = TRUE)

plot_sumz_density(titan_clad2, ribbon = FALSE, sumz = TRUE, change_points = TRUE)

plot_sumz_density(titan_tp, sumz = FALSE)
plot_sumz_density(titan_tn, sumz = FALSE)
plot_sumz_density(titan_clad, sumz = FALSE)

plot_sumz_density(titan_clad2, sumz = FALSE)


titan_tp$sumz|>
round(1)
titan_tp2$sumz|>
round(1)

titan_tn$sumz|>
round(1)
titan_tn2$sumz|>
round(1)

titan_clad$sumz|>
  round(3)
titan_clad2$sumz|>
  round(3)
```
